---
title: 'Ch 3: Sampling the Imaginary'
output:
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo=FALSE, include=FALSE}
library(cowplot)
library(kableExtra)
library(knitr)
library(tidyverse)
```

What's the probability of someone being a vampire?

```{r 3.1}
PrPV <- 0.95
PrPM <- 0.01
PrV <- 0.001

PrP <- PrPV * PrV + PrPM * (1 - PrV)
PrVP <- PrPV * PrV / PrP
PrVP
```

## Sampling from a grid-approximate posterior

Generate 1,000 values for the posterior distribution (back to the globe example from Ch 2)

```{r 3.2}
p_grid <- seq(0, 1, length.out = 1e3)
prior <- rep(1, length(p_grid))
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- (likelihood * prior) %>% {. / sum(.)}
```

Draw 10,000 samples from the posterior

```{r 3.3}
samples <- sample(p_grid, prob = posterior, size = 1e4, replace= TRUE)
```

Plot the samples and their density

```{r 3.4}
sample_tbl <- tibble(i = seq_along(samples), sample = samples)
plot_grid(
  ggplot(sample_tbl, aes(i, sample)) +
    geom_point(color = "blue", fill = "light blue", alpha = 0.5, shape = 21) +
    expand_limits(y = c(0, 1)) +
    theme_classic(),
  ggplot(sample_tbl, aes(sample)) +
    geom_density() +
    expand_limits(x = c(0, 1)) +
    theme_classic()
)
```

## Sampling to summarize

### Intervals of defined boundaries

Posterior probability the proportion of water is less than 0.5?

```{r 3.6}
# add up the posterior probability where p < 0.5
sum(posterior[p_grid < 0.5])
```

Now with the samples

```{r 3.7}
# below 0.5?
sum(samples < 0.5) / length(samples)
# between 0.5 and 0.75?
sum(samples > 0.5 & samples < 0.75) / length(samples)
```

### Intervals of defined mass

Where's the 80th percentile?

```{r 3.9}
quantile(samples, 0.8)
```

What's the middle 80%?

```{r 3.10}
quantile(samples, c(0.1, 0.9))
```

How about with a skewed distribution? The 50% **percentile interval** no longer contains the mode of the distribution and is therefore misleading.

```{r 3.11}
p_grid <- seq(0, 1, length.out = 1e3)
prior <- rep(1, length(p_grid))
likelihood <- dbinom(3, size = 3, prob = p_grid)
posterior <- (likelihood * prior) %>% {. / sum(.)}
samples <- sample(p_grid, size = 1e4, replace = TRUE, prob = posterior)
PI(samples, prob = 0.5)
```

Conversely, the **highest posterior density interval** is the narrowest interval containing the specified probability mass.

```{r 3.13}
HPDI(samples, prob = 0.5)
```

### Point estimates

**The entire posterior distribution is the Bayesian parameter estimate**, but if you have to report one number what should you choose?

Mode, mean, median?

```{r 3.14}
# maximum a posteriori estimate
p_grid[which.max(posterior)]

# mode of the samples
dsamples <- density(samples, adj = 0.01)
dsamples$x[which.max(dsamples$y)]

# other central tendencies
mean(samples)
median(samples)
```

## Sampling to simulate prediction

Generate samples from a binomial and check their densities.

```{r 3.23}
dummy_w <- rbinom(1e5, size = 2, prob = 0.7)
table(dummy_w) / length(dummy_w)

dbinom(0:2, size = 2, prob = 0.7)
```

**Posterior predictive distribution** - what does a parameter predict? It's not the same as the posterior distribution itself, because each $p$ generates a different distribution of predictions. But we can average those prediction distributions of the posterior distribution.

```{r 3.25}
# This is the prediction distribution for a given p
w <- rbinom(1e4, size = 9, prob = 0.6)
simplehist(w)

# This is the prediction distribution averaged over all p's
p_grid <- seq(0, 1, length.out = 1e3)
prior <- rep(1, length(p_grid))
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- (likelihood * prior) %>% {. / sum(.)}
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)
w <- rbinom(1e4, size = 9, prob = samples)
simplehist(w)
```

## Exercises

### Easy

```{r 3.27}
p_grid <- seq(0, 1, length.out = 1e3)
prior <- rep(1, length(p_grid))
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- (likelihood * prior) %>% {. / sum(.)}
set.seed(100)
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)
```

3e1

```{r 3e1}
sum(samples < 0.2) / length(samples)
```

3e2

```{r 3e2}
sum(samples > 0.8) / length(samples)
```

3e3

```{r 3e3}
sum(samples > 0.2 & samples < 0.8) / length(samples)
```

3e4

```{r 3e4}
quantile(samples, 0.2)
```

3e5

```{r 3e5}
quantile(samples, 0.8)
```

3e6

```{r 3e6}
HPDI(samples, prob = 0.66)
```

3e7

```{r 3e7}
quantile(samples, c((1 - 0.66) / 2, 1 - (1 - 0.66) / 2))
```

### Medium

3m1 

```{r 3m1}
p_grid <- seq(0, 1, length.out = 1e3)
prior <- rep(1, length(p_grid))
likelihood <- dbinom(8, size = 15, prob = p_grid)
posterior <- (likelihood * prior) %>% {. / sum(.)}
```

3m2

```{r 3m2}
samples <- sample(p_grid, 1e4, prob = posterior, replace = TRUE)
HPDI(samples, prob = 0.9)
```

3m3

```{r 3m3}
w <- rbinom(1e4, size = 15, prob = samples) 
simplehist(w)
table(w) / length(w)
```

Probability of 8 water in 15 tosses is 14.1%.

3m4

Probability of 6 water in 15 tosses is 11.7%

3m5

```{r 3m5}
# with step prior
p_grid <- seq(0, 1, length.out = 1e3)
prior <- ifelse(p_grid < 0.5, 0, 1)
likelihood <- dbinom(8, size = 15, prob = p_grid)
posterior <- (likelihood * prior) %>% {. / sum(.)}

# repeat earlier problems
HPDI(samples, prob = 0.9)
samples <- sample(p_grid, 1e4, prob = posterior, replace = TRUE)
w <- rbinom(1e4, size = 15, prob = samples) 
table(w) / length(w)

# compare to p = 0.7 (true value)
dbinom(c(6, 8), 15, 0.7 )
```

90% HPDI shifted from (0.33, 0.72) to (0.50, 0.711). Probability of 8 in 15 shifted from 14.1% to 16.1% (compared to 8.11% at p = 0.7) and 6 in 15 from 11.7% to 6.91% (compared to 1.16% at p = 0.7).

### Hard

```{r 3.29}
data(homeworkch3)
# birth1 and birth2
```

3h1

```{r 3h1}
p_grid <- seq(0, 1, length.out = 1e3)
prior <- rep(1, length(p_grid))
likelihood <- dbinom(sum(c(birth1, birth2)), 
                     size = length(c(birth1, birth2)), 
                     prob = p_grid)
posterior <- (likelihood * prior) %>% {. / sum(.)}
p_grid[[which.max(posterior)]]
```

3h2 

```{r 3h2}
samples <- sample(p_grid, 1e4, replace = TRUE, prob = posterior)
HPDI(samples, prob = c(0.5, 0.89, 0.97))
```

3h3

```{r 3h3}
simbirth <- rbinom(1e4, size = 200, prob = samples)
dens(simbirth)
abline(v = sum(birth1, birth2))
```

3h4

```{r 3h4}
simbirth <- rbinom(1e4, size = 100, prob = samples)
dens(simbirth, adj = 0.6)
abline(v = sum(birth1))
```

Model less accurate on first births

3h5

```{r 3h5}
simbirth <- rbinom(1e4, size = sum(!birth1), prob = samples)
dens(simbirth, adj = 1)
abline(v = sum(birth2[!birth1]))
```

The second child is more likely to be a boy following the birth of a girl.
