---
title: 'Ch 5: Linear Models'
output:
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

## Spurious association

Correlation between marriage and divorce

```{r 5.1}
library(cowplot)
library(rethinking)
library(tidyverse)

# data
data(WaffleDivorce)
z <- function(x) (x - mean(x)) / sd(x)
d <- WaffleDivorce %>% 
  mutate(MedianAgeMarriage.s = z(MedianAgeMarriage),
         Marriage.s = z(Marriage))

# model
m5.1 <- rethinking::map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bA * MedianAgeMarriage.s,
    a ~ dnorm(10, 10),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

# plot
MAM.seq <- seq(-3, 3.5, length.out = 30)
mu <- link(m5.1, data = data.frame(MedianAgeMarriage.s = MAM.seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

pred_tbl <- tibble(
  MedianAgeMarriage.s = MAM.seq,
  Divorce = mu.mean,
  Divorce_low = mu.PI[1, ],
  Divorce_high = mu.PI[2, ]
)

ggplot(d, aes(MedianAgeMarriage.s, Divorce)) +
  geom_point() +
  geom_ribbon(aes(ymin = Divorce_low, ymax = Divorce_high),
              pred_tbl,
              alpha = 0.2) +
  geom_line(data = pred_tbl) +
  theme_classic() + 
  theme(aspect.ratio = 1)
```

```{r 5.3}
# model
m5.2 <- rethinking::map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bR * Marriage.s,
    a ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

# plot
mar.seq <- seq(-2, 3, length.out = 30)
mu <- link(m5.2, data = data.frame(Marriage.s = mar.seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

pred_tbl <- tibble(
  Marriage.s = mar.seq,
  Divorce = mu.mean,
  Divorce_low = mu.PI[1, ],
  Divorce_high = mu.PI[2, ]
)

ggplot(d, aes(Marriage.s, Divorce)) +
  geom_point() +
  geom_ribbon(aes(ymin = Divorce_low, ymax = Divorce_high),
              pred_tbl,
              alpha = 0.2) +
  geom_line(data = pred_tbl) +
  theme_classic() + 
  theme(aspect.ratio = 1)
```

Specifying a multivariate model
$$
D_i \sim \mathcal{N}(\mu_i, \sigma) \\
\mu_i = \alpha + \beta_R R_i + \beta_A A_i \\
\alpha \sim \mathcal{N}(10, 10) \\
\beta_R \sim \mathcal{N}(0, 1) \\
\beta_A \sim \mathcal{N}(0, 1) \\
\sigma \sim \mathcal{U}(0, 10)
$$

```{r 5.4}
m5.3 <- rethinking::map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bR * Marriage.s + bA * MedianAgeMarriage.s,
    a ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.3)

plot(precis(m5.3))
```

### Plotting multivariate posteriors

#### Predictor residual plots

First, the residual of one predictor w.r.t. the other

```{r 5.6}
m5.4 <- rethinking::map(
  alist(
    Marriage.s ~ dnorm(mu, sigma),
    mu <- a + b * MedianAgeMarriage.s,
    a ~ dnorm(0, 10),
    b ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

# compute expected value at MAP, for each State
mu <- coef(m5.4)['a'] + coef(m5.4)['b'] * d$MedianAgeMarriage.s
# compute residuals
m.resid <- d$Marriage.s - mu

# plot residuals
R_resid_tbl <- tibble(
  id = rep(d$Location, 2),
  MedianAgeMarriage.s = rep(d$MedianAgeMarriage.s, 2),
  Marriage.s = c(mu, d$Marriage.s)
)
mu_tbl <- tibble(
  MedianAgeMarriage.s = range(d$MedianAgeMarriage.s),
  Marriage.s = coef(m5.4)['a'] + coef(m5.4)['b'] * MedianAgeMarriage.s
)
ggplot(d, aes(MedianAgeMarriage.s, Marriage.s)) +
  geom_line(data = mu_tbl) +
  geom_line(aes(group = id), R_resid_tbl) +
  geom_point(color = "blue", shape = 21) +
  theme_classic() +
  theme(aspect.ratio = 1)
```

#### Counterfactual plots

```{r 5.9}
# prepare counterfactual data
A.avg <- mean(d$MedianAgeMarriage.s)
R.seq <- seq(-3, 3, length.out = 30)
pred.data <- tibble(
  Marriage.s = R.seq,
  MedianAgeMarriage.s = A.avg
)

# compute counterfactual mean divorce (mu)
mu <- link(m5.3, data = pred.data)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# simulate counterfactual divorce outcomes
R.sim <- sim(m5.3, data = pred.data, n = 1e4)
R.PI <- apply(R.sim, 2, PI)

# display predictions, hiding raw data with type = "n"
plot(Divorce ~ Marriage.s, data = d, type = "n")
mtext("MedianAgeMarriage.s = 0")
lines(R.seq, mu.mean)
shade(mu.PI, R.seq)
shade(R.PI, R.seq)
```

#### Posterior prediction plots

```{r 5.11}
#call link without specifying new data so it uses original data
mu <- link(m5.3)

# summarize samples across cases
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# simulate observations
# again no new data, so uses original data
divorce.sim <- sim(m5.3, n = 1e4)
divorce.PI <- apply(divorce.sim, 2, PI)

post_pred <- tibble(
  state = d$Location,
  divorce = d$Divorce,
  mu.mean,
  mu.low = mu.PI[1, ],
  mu.hi = mu.PI[2, ],
  divorce.low = divorce.PI[1, ],
  divorce.hi = divorce.PI[2, ],
  resid = divorce - mu.mean,
  resid.low = divorce - divorce.low,
  resid.hi = divorce - divorce.hi
)

ggplot(post_pred, aes(divorce, mu.mean, ymin = mu.low, ymax = mu.hi)) +
  geom_pointrange(shape = 21, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Observed divorce", 
       y = "Predicted divorce") +
  theme_classic() +
  theme(aspect.ratio = 1)
```

Now plot error for each state

```{r 5.14}
post_pred %>% 
  mutate(state = fct_reorder(factor(state), resid)) %>% 
  ggplot(aes(state, resid, ymin = resid.low, ymax = resid.hi)) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  geom_pointrange(shape = 21) +
  coord_flip() +
  labs(x = "", y = "Divorce residual") +
  theme_classic()
```

Spurious data example. Note the 89% PI for $b_{real}$ overlaps 1 and for $B_{spur}$ overlaps 0.

```{r 5.15}
N <- 1e2
x_real <- rnorm(N)
x_spur <- rnorm(N, x_real)
y <- rnorm(N, x_real)
d <- tibble(x_real, x_spur, y)
pairs(d)

m.spur <- rethinking::map(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- a + bR * x_real + bS * x_spur,
    a ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    bS ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m.spur)
```

## Masked relationship

```{r 5.16}
data(milk)
d <- milk %>% 
  drop_na(neocortex.perc)
str(d)

m5.5 <- rethinking::map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = d
)

precis(m5.5, digits = 3)

np.seq <- 0:100

mu <- link(m5.5, data = tibble(neocortex.perc = np.seq), n = 1e4)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

pred.data <- tibble(
  neocortex.perc = np.seq,
  kcal.per.g = mu.mean,
  kcal.per.g.low = mu.PI[1, ],
  kcal.per.g.hi = mu.PI[2, ]
)

ggplot(d, aes(neocortex.perc, kcal.per.g)) +
  geom_point(shape = 21, color = "blue") +
  geom_line(data = pred.data) +
  geom_ribbon(aes(ymin = kcal.per.g.low, ymax = kcal.per.g.hi),
              pred.data,
              alpha = 0.3)
```

```{r 5.24}
d2 <- mutate(d, log.mass = log(mass))

m5.6 <- rethinking::map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bm * log.mass,
    a ~ dnorm(0, 100),
    bm ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = d2
)
precis(m5.6)
```

So we see two variables with weakly contrasting effects: kcal per g barely increases with neocortex percentage and barely decreases with log mass. So the two variables are masking each other's influence and bivariate regression is misleading. A multivariate regression revelas how strong the effect actually is.

```{r 5.26}
m5.7 <- rethinking::map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc + bm * log.mass,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    bm ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = d2
)

precis(m5.7)

# Vary neocortex percent but keep log.mass fixed
mean.log.mass <- mean(log(d2$mass))
np.seq <- 0:100
pred.data1 <- data.frame(
  neocortex.perc = np.seq,
  log.mass = mean.log.mass
)

mu <- link(m5.7, data = pred.data1, n = 1e4)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

pred.data1$kcal.per.g <- mu.mean
pred.data1$kcal.per.g.low <- mu.PI[1, ]
pred.data1$kcal.per.g.hi <- mu.PI[2, ]

# And vice versa
mean.np <- mean(d2$neocortex.perc)
log.mass.seq <- seq(-2.5, 4.5, length.out = 100)
pred.data2 <- data.frame(
  neocortex.perc = mean.np,
  log.mass = log.mass.seq
)

mu <- link(m5.7, data = pred.data2, n = 1e4)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

pred.data2$kcal.per.g <- mu.mean
pred.data2$kcal.per.g.low <- mu.PI[1, ]
pred.data2$kcal.per.g.hi <- mu.PI[2, ]


plot_grid(
  ggplot(d2, aes(neocortex.perc, kcal.per.g), d2) +
    geom_point(shape = 21, color = "blue") +
    geom_line(data = pred.data1) +
    geom_ribbon(aes(ymin = kcal.per.g.low, ymax = kcal.per.g.hi),
                pred.data1,
                alpha = 0.2) +
    theme_classic() +
    theme(aspect.ratio = 1),
  ggplot(d2, aes(log.mass, kcal.per.g), d2) +
    geom_point(shape = 21, color = "blue") +
    geom_line(data = pred.data2) +
    geom_ribbon(aes(ymin = kcal.per.g.low, ymax = kcal.per.g.hi),
                pred.data2,
                alpha = 0.2) +
    theme_classic() +
    theme(aspect.ratio = 1),
  nrow = 1
)
```

## When adding variables hurts

### Multicollinear legs

```{r 5.29}
N <- 100

d <- tibble(
  height = rnorm(N, 10, 2),
  leg_prop = runif(N, 0.4, 0.5),
  leg_left = leg_prop * height + rnorm(N, 0, 0.02),
  leg_right = leg_prop * height + rnorm(N, 0, 0.02)
)

m5.8 <- rethinking::map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl * leg_left + br * leg_right,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    br ~ dnorm(2, 10),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

precis(m5.8)

plot(precis(m5.8))

post <- extract.samples(m5.8)
plot(bl ~ br, post, col = col.alpha(rangi2, 0.1), pch = 16)

sum_blbr <- post$bl + post$br
dens(sum_blbr, col = rangi2, lwd = 2, xlab = "sum of bl and br")

m5.9 <- rethinking::map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl * leg_left,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

precis(m5.9)
```

### Multicollinear milk

```{r 5.35}
d <- milk
#kcal.per.g regressed on perc.fat
m5.10 <- rethinking::map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bf * perc.fat,
    a ~ dnorm(0.6, 10),
    bf ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
# kcal.per.g regressed on perc.lactose
m5.11 <- rethinking::map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bl * perc.lactose,
    a ~ dnorm(0.6, 10),
    bl ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

precis(m5.10, digits = 3)
precis(m5.11, digits = 3)
```

Kcal per gram of milk has a postive slope w.r.t. percent fat and negative w.r.t. percent lactose. However in a multivariate regression they cancel out because fat and lactose correlate so strongly.

```{r 5.37}
m5.12 <- rethinking::map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bf * perc.fat + bl * perc.lactose,
    a ~ dnorm(0.6, 10),
    bf ~ dnorm(0, 1),
    bl ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.12, digits = 3)

pairs(~ kcal.per.g + perc.fat + perc.lactose, data = d, col = rangi2)

cor(d$perc.fat, d$perc.lactose)
```

### Post-treatment bias

```{r 5.41}
# number of plants
N <- 100

# simulate initial heights
h0 <- rnorm(N, 10, 2)

# assign treatments and simulate fungus and growth
treatment <- rep(0:1, each = N / 2)
fungus <- rbinom(N, size = 1, prob = 0.5 - treatment * 0.4)
h1 <- h0 + rnorm(N, 5 - 3 * fungus)

d <- data.frame(h0, h1, treatment, fungus)
```

Including both treatment and presence of fungus, the coefficient for treatment is essentially 0 because it acts through controlling fungus. This model asks the wrong question. 

```{R 5.42}
m5.13 <- rethinking::map(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- a + bh * h0 + bt * treatment + bf * fungus,
    a ~ dnorm(0, 100),
    c(bh, bt, bf) ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

precis(m5.13)
plot(precis(m5.13))
```

Leaving out presence of fungus reveals the treatment had a strong effect.

```{R 5.43}
m5.14 <- rethinking::map(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- a + bh * h0 + bt * treatment,
    a ~ dnorm(0, 100),
    c(bh, bt) ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

precis(m5.14)
plot(precis(m5.14))
```

## Categorical variables






